---
title: "Expressions with Rlang"
author: "Kirsten"
date: "25 sep 2018"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library("rlang")
library("lobstr")
```

## Install needed packages
```{r,eval=FALSE,echo=TRUE}
devtools::install_github("r-lib/rlang")
devtools::install_githu("hadley/lobstr") 
# Rtools version 3.4 are needed for lobstr
require("rlang","lobstr")
```

## Introduction I

Save expressitions not containg values
```{r,error=TRUE,echo=TRUE}
y <- x * 10

z <- rlang::expr(y <- x * 10)
z
```
So we see, that `expr()` add quotationmarks. To evaluate the expression for a value of x, we can write

## Introduction II
```{r,echo=TRUE,results=TRUE}
x <- 10
l <- eval(z) 
l
```
Notice that `eval()`is a scilent function so we need to assign the output to a value.

## AST

We have just created what we call a \textbf{quoted expression}. These are what we call \textbf{abstract syntax trees} (AST).

Illustration on blackboard for a function `f(x,"y",1)`. 

In AST we define.

 - \textbf{Calls}, define the hierarchy of the tree.  This is mainly relevant when we consider multiple functions.
 - \textbf{The children}, which are ordered as following: \textbf{The first child} are the function, \textbf{the subsequent children} are the arguments. 

As we have learned earlier the order of the children are important.

## AST II
- \textbf{The leaves} (another name for the children), are \textbf{symbols}, that is objects without a value, such as `f`and `x` or \textbf{constants} such as `1` or `"Y"`. 

## AST III
Show the AST tree with `lobstr::ast()`
```{r, echo=TRUE}
lobstr::ast(f(x,"y",1))
```

Q: Run this code in R

## Unquoting 

We can unquote an expression using the function `!!`. 
Try to Write the following in R, What would you expect, what do you see?

```{r,echo=TRUE,eval=FALSE}
x <- expr(f(x,"y",1))

lobstr::ast(x)
lobstr::ast(!!x)
```

## Unquoting II
```{r,echo=TRUE}
x <- rlang::expr(f(x,"y",1))

lobstr::ast(x)
lobstr::ast(!!x)
```

## Infix and prefix
Try to run the following code.
```{R,eval=FALSE,echo=TRUE}
View(expr(y <- x*10))
```
You will get a detailed describtion, containing, the type of the children, the evaluation order, and the functions. You can even go through the AST tree by unfolding the different calls.

## Infix and prefix
 
 - \textbf{Infix form} is when a function come inbetween arguments, such as `<-` and `*` above. 
 - \textbf{Prefix functions} are functions where the arguments come first. 
 
An infix function can be turned into a prefix function 
```{r, echo=TRUE}
x <- 4
y <- x * 10
y
`<-`(z, `*`(x, 10))
z
```

Draw the AST on the blackboard.

## Exercise 

Print the AST for `names(x)<-y`.

- Which calls to you make?
- What type are the children, and what is the order of the children?

Try to run `lobstr::ast(mean(x = mtcars$cyl, na.rm = TRUE))` (you might need to type `data(mtcars)` first.). 

 - Can you find a infix and a prefix function.
 - Is the output as expected?
 
Other special forms that we can look at are `for()`, `if()`, `[`, `[[`,`{` and many more.

## Grammar

The rules that governer the trees constructed by a sequence of tokens, are called the \textbf{grammar}.

This includes the order of which operations ar processed, forexample do R read `(1+2)*3` or `1+(2*3)`, if we write `1+2*3`.

Q: What would you expect?

- Write `lobstr::ast(1 + 2 * 3)`to find out.
- What if we write `lobstr::ast(!x %in% y)`?

If you write `lobstr::ast(1 + 2 + 3)` you see, that R is \textbf{left-associative}, meaning values on the left are evaluated first. However the exceptions that confirms the rule are `<-` and `^`. We can sometimes overrule the left associative. To see this try to consider `x + y %+% z` and `x ^ y %+% z`.

## Data structure

\textbf{Expressions} are used to refer to the set containing \textbf{constants, symbols, pairlists} and \textbf{calls}. 

In \textbf{base R} expressions (called \textbf{expressions object} in [HW]) is a special type equivalent to a list of expressions.

The expressions we will be working with are closest to the R base object called \textbf{langauge object} (including symbols and calls). The type you saw ealier when using `view()`.

## Constants

\textbf{Constants} occure in the leaves of AST, and are the simplest datastructure found in the AST (that is they are \textbf{atomic vectors} of length 1).

Constants are self quoting since 
```{r,echo=TRUE}
identical(expr("x"), "x")
identical(expr(TRUE), TRUE)
```

Q: do you think the same goes for `identical(expr(1), 1)`? Try to type it.

## Symbols I

\textbf{Symbols} represent variable names. We can convert back and fourth between symbols and the string that represents them.
```{r,echo=TRUE}
"x"
sym("x")
as_string(sym("x"))
```

## Symbols II

We can also use syms to put the symbols in a list. 
```{r,echo=TRUE}
syms(c("h","hey"))
```

## Symbols III

Symbols differ from string mainly in evaluation. 
```{r,echo=TRUE,error=TRUE}
x <- 10
y <- "5"

z<- eval(expr(f <- x*4))
z
y <- eval(expr(f <- y*4))
```

## Missing arguments

A special symbol is the missing arguments. 
```{r,echo=TRUE}
rlang::missing_arg()
rlang::type_of(missing_arg())
rlang::as_string(missing_arg())
```

Notice that the type of `missing_arg()` is a symbol, but if we turn it into a string it contains no value. 

## Mising arguments II

To see if a symbol is missing, we can use 
```{r,echo=TRUE}
rlang::is_missing(missing_arg())
```

We can bind it to a variable `m <- missing_arg()`, what would you expect the output to be?

## Missing arguments III

```{r,error=TRUE, echo=TRUE}
m <- missing_arg()
m
```

If we instead write 
```{r,echo=TRUE}
mm <- list(missing_arg())
mm[[1]]
```
We will not get an error or an output at all. 

To be able to work with missing arguments stored in a variable we can write,
```{r, echo=TRUE}
rlang::maybe_missing(m)
```

## Calls

\textbf{Calls} define the AST, and behave similar to a list. 
```{r,echo=TRUE}
x <- expr(read.table("important.csv", row = FALSE))

ast(!!x)
```

## Calls II
```{r,echo=TRUE}
length(x) -1 # Get the number of arguments
names(x) # Missing arguments
c(x[[1]],x[[2]]) # Extract leaves
x$row
```

## Extract arguments I

```{r, echo=TRUE}
rlang::call_standardise(x)
```

`call_standardise()` standardises all arguments to use the full name. However if the function uses `...` it is not possible to standardise all arguments. 

## Extract arguments II

Now we can extract multiple arguments 
```{r,echo=TRUE}
as.list(x[2:3])
```
It is possible to use without `as.list()`, but it is considererd good pratice.

## Modify calls 

```{r,echo=TRUE}
x$header <- TRUE
x
```


## Construct calls

Finaly, we can construct a call from its children using `call2()`

```{r,echo=TRUE}
call2("mean", x = expr(x), na.rm = TRUE)
call2(expr(mean), x = expr(x), na.rm = TRUE)

x <- 1:10
eval(call2(expr(mean), x = expr(x), na.rm = TRUE))
```

Q: try to construct a call using your favorite function. 

## Pairlists
\textbf{Pairlists} have mostly been replaced by lists, but will occure when working with function arguments. 

```{r,echo=TRUE}
f <- function(x = 10) x + 1
typeof(formals(f))
```

The disadvantage is that the datastructure is a linked list instrad of a vector, so subsetting gets slower the further down the pairlist you index.

## Pairlists II

In C the pairlists are still used. But we can treat them as a list.

```{r,echo=TRUE}
pl <- pairlist(x = 1, y = 2)
length(pl)
str(pl)
```

To illustrate the disadvantage

```{r, echo=TRUE}
l1 <- as.list(1:100)
l2 <- as.pairlist(1:100)
```

## Pairlist III

`supressWarnings()` not to print the warning that the computation time of `l1[[1]]`and `l1[[100]]` are to short to measure

```{r,echo=TRUE}
suppressWarnings(microbenchmark::microbenchmark(
  l1[[1]],
  l1[[100]],
  l2[[1]],
  l2[[100]]))
```

## Expression objects. 

Remember these are a equivalent to a list of expressions. 

We can produce an expression object in the two following ways in base R.
```{R,echo=TRUE}
exp1 <- parse(text = c("
x <- 4
x
"))
exp2 <- expression(x <- 4,x)

typeof(exp1)
typeof(exp2)
```

## Expression objects II

Q: Try to write `length(exp1)` and `exp1[[1]]`.

If we use `eval()` on a expression object, we evaluate all expressions and thus get a list of expressions. 

## Parsing and deparsing

How can we evaluate an expression if it is stored as a string, lets say 

```{r,echo=TRUE}
x <- "y <- z + 7"
lobstr::ast(!!x)
```

## Parsing I
Using `rlang()` we can do the following

```{r,echo=TRUE}
x1 <- rlang::parse_expr(x)
x1

# With AST
lobstr::ast(!!x1)
```

## Parsing II

For multiple expressions in a string 
```{r,echo=TRUE, eval=FALSE}
x <- "a <- 1; a + 1"
rlang::parse_exprs(x)
```
Q: What would you expect?

## Parsing III

```{r,echo=TRUE}
x <- "a <- 1; a + 1"
rlang::parse_exprs(x)
```

The basse R function equivalent to `parse_exprs()` is `parse()`.

## Deparsing I

```{r, echo=TRUE}
z <- expr(y <- x + 10)
expr_text(z)
```

With R base equivalent function being `deparse()`.

## Quasiquotation

\textbf{Quasiquotation} is one of the fundamental ideas that make the functions `expr()` and `ast()` work. 

 - \textbf{quotation} allows us to capture AST associated with an argument.
 
 - \textbf{Unquotation} allows you to selectively evaluate parts of an quoted expression.
 
## Quotation I

Why do we need unquoting?

```{r,echo=TRUE}
paste("It", "takes", "a","while","to","type","all","these","words")
```

## `exprs()` I

The function`exprs()`makes this easier
```{r,echo=TRUE}
cement <- function(...) {
  dots <- exprs(...)
  paste(purrr::map(dots, expr_name), collapse = " ")
}

cement(now,I,can,write,the,words,without,quotation)
```

## `exprs()` II

However it doesnt solve all problems

```{r,echo=TRUE}
day <- "Friday"

paste("Today is",day)
cement("Today is",day)
```

Q: Can you remember a function that might help us?

## Unquoting: Introduction

To solve this problem we need a function to unqoute.
```{r,echo=TRUE}
cement("Today is",!!day)
```

## With rlang
4 functions, 

 - `expr()` and `enexpr()`
 - `exprs()` and `enexprs()` 

We have already used `expr()`.
```{r,echo=TRUE}
expr(1 /2 / 3)
```

## Exercise
What would you expect the output from the following code to be?
```{r,echo=TRUE,eval=FALSE}
f1 <- function(x) expr(x)
f1(a + b + c)
```


## Answer
```{r,echo=TRUE}
f1 <- function(x) expr(x)
f1(a + b + c)
```

## `enexpr()`
But with `enexpr()`
```{r,echo=TRUE}
f1 <- function(x) enexpr(x)
f1(a + b + c)
```

## `expres()`

With `exprs()` we can evaluate a list
```{r,echo=TRUE}
exprs(x = x ^ 2, y = y ^ 3, z = z ^ 4)
```

## `expres()` II
The function also returns missing arguments
```{r,echo=TRUE}
val <- exprs(x = )
is_missing(val$x)
```

In conclusion use `ènexpr()` and `enexprs()` inside functions and `expr()` and `exprs()` to capture expressions. 

## Exercise

What does teh following code return, and why?
```{r,echo=TRUE,eval=FALSE}
expr({
  x +              y # comment  
})
```

## Answer
```{r,echo=TRUE}
expr({
  x +              y # comment  
})
```

## Unquoting: rlang
To unquote with rlang we will use the function `!!`. 

We have seen before
```{r,echo=TRUE}
x <- expr(a + b + c)
expr(f(!!x, y))
```

Diagram of AST?

## Unquoting: rlang II

What happens if we write
```{r,echo=TRUE}
x <- exprs(1, 2, 3, y = 10)
expr(f(!!x, z = z))
expr(f(!!!x, z = z))
```
 !!! takes a list of expressions and insert them at the location of `!!!`.
 
## Problems with `!!!`

```{r,echo=TRUE}
x <- 100
head(mtcars$cyl,7)
head(with(mtcars, cyl + !!x),7)
head(with(mtcars, cyl + !x),7)
```

Q: What woud you expect to get if we wrote `!!!x` instea?

## Additional
`rlang::eval_bare()` is another relevant function if you want to evaluate an expression.